Synthesis of Rules for Network Intrusion Detectors

### 2019-10-23 paper/paper.tex
pass

This paper focuses on signature-based \nids, Suricata, in
particular. But the principles we use are general to any other
signature-based \nids~(\eg{}, Snort~\cite{snort}).

@@

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}. A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied. In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Some options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') whereas the remaining properties
describe characteristics of the attack (\eg, ``flags'', ``threshold'',
and ``content''). The main challenge in synthesizing rules is to
discover the set of options that make a rule to capture the malicious
traffic, but ignore the benign traffic.

\Fix{tabulate some important rules that we support}

@@

\subsection{SYN Flood Attack}

The SYN Flood attack exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.
Figure~\ref{fig:synflood-example} shows a Suricata rule to capture
this attack. The critical parts appear in bold---the option
\CodeIn{flags: S,12}, which identifies a SYN packet and the option
\CodeIn{threshold: type both, track by\_dst, count 5000, seconds 5},
indicating that a high volume of such packets should be requested in a
short period of time.

%\Mar{please, explain the other fields}

\subsubsection{\tname\ on SYN Flood Attack}

@@

\Fix{repeat this for two other attacks. one non-dos/flood non-content}

@@

\subsection{Problem Characterization}

One important step in search-based optimization is to characterize a
candidate solution to the target problem. Candidate solutions are the
objects that the search optimize towards solutions. In this case, a 
candidate solution is an encoding of a Suricata rule.  \tname\ encodes
the rule options segment of a Suricata rule (see
Section~\ref{sec:example-suricata-rules}) as a vector of Option
objects. Some options are discarded as they only serve for
documentation. For example, the options \CodeIn{msg},
\CodeIn{classtype}, and \CodeIn{misc-activity}, appearing in the
example on Figure~\ref{fig:synflood-example}, have no influence in
triggering the actions associated with the rule. The option
\CodeIn{flags}, however, is relevant to trigger the action. \tname{}
models that option with a data structure storing two fields, one of
character type and another of integer type. This enables \tname{} to
mutate each element separately. Similar encodings are used in modeling
the other relevant options. We found what options are (ir)relevant by
inspecting Suricata code for the presence of \CodeIn{eval} functions,
which are invoked from a main loop to process each action rule. The
presence of such a function for a given option indicates that it is
relevant for triggering some action of some rule.

\subsection{Initial State}

To bootstrap the search, \tname{} needs to select an initial
solution. When creating such candidate solution, \tname{} first looks
for the protocol related to the malicious input traffic. Then, it
selects the relevant options for that protocol and initializes the
option vector with default values. \Mar{Suricata define valores
  default?  Vcs. definiram?  Isto foi baseado em popularidade ou o
  que?}

%% \tname{}
%% uses an abstract representation of a Suricata rule that reflects the
%% concrete representation of the rule. The concrete rule representation
%% is the one actually used by Suricata whereas the abstract
%% representation is the one \tname{} uses to run the search.

\subsection{Fitness Functions}

The fitness value of a candidate solution $c$, modeling the rule
options $o_1, \dots, o_n$ is given by $f(c)=1-\Sigma{d(o_i,m)}/n$,
where $i$ ranges from 1 to $n$ and $m$ denotes the negative
traffic. The higher the fitness value the better, \ie{}, the fitter
the candidate solution $c$ will be.  The value of the expression
$d(o_i,m)$, associated with option $o_i$, ranges over
0-1. Consequently, the value of the expression $\Sigma{d(o_i,m)}/n$
itself is in the 0-1 range, as there are $n$ options, and the range of
the function is 0-1.

As usual, we used customized distance functions to compute $d(o_i,
m)$. For example, let us consider the \CodeIn{flags} option, discussed
on Figure~\ref{fig:synflood-example}. Recall that this option
identifies the SYN packet in the tcp protocol and that its encoding in
\tname{} stores two attributes---one categorical and one
numerical. The value of $d(o_i, m)$, in this case, is obtained by
computing the normalized Euclidean distance in the two-dimensional
space. Considering that each of the two dimensions has cardinality of
ten, the distance of a candidate solution holding the option
\CodeIn{flags(S,13)} to the solution will be 0.22
(=$1/\sqrt{20}$). Note that the smaller the distance the better.

\tname{} uses a modified version of Suricata to measure distance. It
updates the rule set that the modified version of Suricata analyses,
injects the traffic on the network, and monitors the execution. Each
rule option has an associated \CodeIn{eval} function in code. \tname{}
monitors the execution of these functions to identify how far it is
from returning true, which indicates it is closer to trigger the
action associated with the rule. 
\Luc{Originalmente, o suricata interrompe a checagem de uma regra quando alguma das opções não dá match com o pacote analisado. O código fonte foi alterado para que as regras sempre tenham todas as suas opções verificadas. A função de verificação de uma opção retorna um booleano indicando se o valor da opção na regra corresponde ou não ao valor da opção no pacote. Para opções cujos valores são numéricos, alteramos o código fonte para obter a diferença entre os valores da opção na regra e no pacote.
\\*
Exemplo trecho de código original opção "icode":

\begin{figure}[h]
  \lstinputlisting[language=C,numbers=none,keywords={}]{original.code}
  \caption{Original source code.}
  \label{fig:original-code}
\end{figure}

Trecho modificado:

\begin{figure}[h]
  \lstinputlisting[language=C,numbers=none,keywords={}]{modified.code}
  \caption{Modified source code.}
  \label{fig:modified-code}
\end{figure}

\Fix{Lucas/Guilherme, it would be good to show how this looks like in
  code}

\subsection{Search}

\tname{} is a single-individual hill-climbing like search.

>>>>

This paper focuses on signature-based \nids. Although we focused on
the Suricata \nids, the principles we used are general to any other
signature-based \nids~(\eg{}, Snort~\cite{snort}).

@@

Figure~\ref{fig:synflood-example} shows an example rule of
Suricata~\cite{suricata}, a popular open-source \nids{} maintained by
the Open Information Security Foundation (OISF)~\cite{oisf}. A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied. In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Options serve to 1) document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to 2) describe
characteristics of the attack (\eg, ``flags'', ``threshold'', and
``content''). \tname{} produces options of the second kind, which are
\emph{relevant} to capturing the attack.

Suricata supports a total of \totoptions\ options covering
\numproto\ distinct protocols. Of these total of options,
\totoptionsrelevant\ are relevant to capture traffic (see
Figure~\ref{fig:distribution-rules-protocol}). Table~\ref{table:rules}
shows a sample of the relevant options of Suricata. Column ``\#''
shows the option number, column ``Name'' shows the name of the
option, column ``\#~Params'' shows the number of additional parameters
associated with a given option, and column ``Description'' presents a
short description of the option purpose.

\begin{table}[h]
  \caption{\label{table:rules}Example options.}  
  \centering
  \begin{tabular}{p{0.5cm}lrp{8cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &
    \multicolumn{1}{c}{\# Params} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & - & matches on the size of the packet payload\\
    2 & itype & - & matches on a specific ICMP type\\
    3 & icode & - & matches on a specific ICMP code\\
    4 & icmp\_seq & - & checks a ICMP sequence number\\
    5 & icmp\_id & - & matches on specific ICMP id-values\\
    6 & window & - & checks the size of the TCP window\\
    7 & flags & - & matches on TCP flags\\
    8 & fragbits & - & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & - & controls the rule’s alert frequency\\
    10 & content & - & matches on bytes of the payload\\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Rules and Packets}

\Fix{Illustrate how information is extracted from packets}

@@

\subsection{Active Reconnaissance}

Active reconnaissance is a method used by malicious individuals to
gather information about vulnerabilities from a computing system. The
most common approach is to ``scan'' open ports to discover
vulnerabilities. Active reconnaissance is a preliminary step towards
an attack that actually exploits the system. A Ping Scan is a very
common type of port scan based on the ICMP
protocol. \Fix{...explain this attack...}

\Fix{...explain how \tname\ works on it...}

\subsection{Denial-of-Service}

The SYN Flood attack is a denial-of-service attack that exploits a vulnerability in the TCP/IP handshake
to establish a TCP connection~\cite{cloudfare-synflood}. The handshake
works as follows in normal circumstances. First, a client sends a SYN
packet to the server, requesting a connection. Second, the server
responds with a SYN-ACK packet to the client. Third, the client
responds with an ACK message and the connection is established. Aware
of the protocol, an attacker sends multiple SYN packets to different
ports of a server, often using fake IP addresses. Then, after the
server responds with a SYN-ACK packet, the client keeps sending other
SYN packets to avoid the connection to time out. Without proper
protection, the server accepts these malicious requests and eventually
legitimate requests cannot be satisfied due to resource exhaustion.
Figure~\ref{fig:synflood-example} shows a Suricata rule that captures
this attack. This rule was obtained from the \Fix{...what source...}
The relevant options in the rule appear in bold---the option
\CodeIn{flags: S,12}, identifies a SYN packet in a TCP packet and the
option \CodeIn{threshold: type both, track by\_dst, count 5000,
  seconds 5} indicates that a high volume of such packets should be
requested in a short period of time.

%% Explain how \tname\ behaves on this attack

The main challenge in synthesizing rules is to discover the set of
options that makes a rule to capture the malicious traffic, but
ignores the benign traffic.

\tname{} synthesizes rules with the action header fixed to
\CodeIn{alert} and with the header \CodeIn{<proto> any any -> any
  any}, indicating that it infers the target protocol and uses the
most general flow description possible, \ie{}, the rule applies to the
traffic flowing from/to any address and port. The rule that \tname{}
synthesizes for the SYN Flood denial-of-service attack is as follows.

\begin{figure}[h!]
  \lstinputlisting[language=C,numbers=none,frame=none,keywords={flags,threshold}]{synflood.suricata.synth}
%  \caption{Suricata rule for SYN Flood Attacks.}
%  \label{fig:synflood-example-synt}
\end{figure}

@@

@@
